Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-03-29 paper/AlignedIsotropicGW.tex
Implemented Ilya's 'red' text.

\begin{abstract}
  The first direct detections of \acp{GW} from merging \acp{BBH} open
  a unique window into the formation environment of massive stars and
  their compact remnants.  One promising signature of the formation
  environment is the angular distribution of the \ac{BH} spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin orientations whereas
  binaries formed from pairs of stars born together are more likely to
  have spins \ilya{preferentially} aligned with the binary orbit as a consequence of their
  joint evolution toward a \ac{BBH} system.  We consider existing
  \ac{GW} measurements of $\chieff$, the best-measured combination of
  spin parameters, in the three likely binary black hole detections
  GW150914, LVT151012, and GW151226.
  \ilya{If binary black hole spin magnitudes extend to high values, as is
suggested by observations of black hole X-ray binaries,} we show that the data already
  exhibit a $\OOneSigmaIsoAligned\sigma$ ($\OOneOddsIsoAligned$ odds
  ratio) preference for an isotropic angular distribution.
  By considering the
  effect of an additional 10 detections drawn from the various models
  in the suite we show that if all observations come from a single
  population such an augmented data set would enable at least a
  $\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds
  ratio) distinction between the isotropic and aligned models \ilya{for the assumed spin magnitude distributions}, and in
  most cases better than $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).
  The existing preference for \ilya{either an isotropic spin distribution or low spin magntiudes} for the observed systems will be confirmed (or overturned) confidently in the near future by subsequent observations.
\end{abstract}

@@

Small values of $\chieff$ can result from either intrinsically small
spins or larger spins whose direction is mis-aligned with the orbital
plane of the binary (i.e.\ spin vectors with small $z$-components).
Mis-alignment is capable of producing \emph{negative} values of
$\chieff$, however, while small, aligned spins will always have
$\chieff \geq 0$.  This difference provides strong discriminating
power between the two angular distributions, even without good
information about the magnitude distribution; to the extent that data
support negative $\chieff$ they weigh heavily against aligned models.
To quantify the degree of support for these two alternate explanations
of small $\chieff$ values in the merging \ac{BBH} population, we
compared the Bayesian evidence for various simple models of the spin
population using the \ac{GW} data set.

@@

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic, which explains why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``decreasing'' magnitude distribution; the most
favoured model among those with an aligned distribution also has the
``decreasing'' magnitude distribution.  The odds ratio between the
aligned and isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
\ac{BBH} systems having zero spin is ruled out by the GW151226
measurements, which bound $\chieff \gtrsim 0.2$ at 90\% credibility
\citep{2016PhRvL.116x1103A}.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the near future
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming an
additional 10 detections drawn from each of our six zero-parameter
models for the spin distribution, we find the odds ratios shown in
Figure \ref{fig:O2-predictions}.  We find that most scenarios with an
additional 10 detections allow the the correct angular distribution to
be distinguished with greater than $5\sigma$ ($2.9 \times 10^{-7}$
odds) credibility; and in the worst case the distinction is
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
Note that such future detections should permit a confident distinction
between \emph{angular} distributions, we will remain very uncertain
about the \emph{magnitude} distribution until we have a much larger
number of observations.  In Figure \ref{fig:O2-predictions}, the odds
ratio between different magnitude distributions with the same angular
distribution is much closer to one than the odds ratio between angular
distributions.

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim \ilya{0.3}$, \ilya{[I changed the number to be consistent with Simon's analysis, which shows that you can no longer tell the difference for a mean spin of $\sim 0.17$]} then
it is no longer possible to resolve the negative spin with a small
number of observations.  As noted below, however, spins observed in
X-ray binaries are typically large. 
\sout{and most models of isolated binary evolution do not produce systems with spins uniformly bounded to small values}.  
Additionally, models which do not permit \emph{some} spins
with $\chieff \gtrsim \ilya{0.1}$ \ilya{[Changed to be consistent with 90\% lower bound from cited paper]} are ruled out by the GW151226 observations
\citep{2016PhRvL.116x1103A}.

In order to perform our analysis we need to select models for the
distribution of the \ilya{spin magnitudes} of stellar-mass
black holes.  Observational data for such model selection is sparse.
\citet{2015PhR...548....1M} give current estimates of the spin
parameters for stellar-mass black holes, obtained using disk
reflection and/or disk continuum methods.  Most of the systems studied
are low-mass X-ray binaries rather than the high-mass X-ray binaries
that are likely to be the progenitors of double black hole binaries.
In addition, there are substantial systematic errors that can
complicate either type of analysis (see \citealt{2015PhR...548....1M}
for a discussion), \ilya{and observational selection effects could yield a biased distribution}. Nonetheless, if we take the reported spin
magnitudes as representative then we find that there is a preference
for high spins; for example, 14 of the 19 systems with reported spins
have dimensionless spin parameters in excess of 0.5.  \ilya{It is usually argued that} the masses and spin parameters
of stellar-mass black holes are unlikely to be altered significantly
by accretion (low-mass donors may not have enough mass and
high-mass donors have a very short phase in which they transfer mass;
a variant of this long-standing argument was presented by
\citet{1999MNRAS.305..654K}), \ilya{but see \citet{Podsiadlowski:2003,FragosMcClintock:2015}}.  Thus the current spin parameters
probably are close to their values upon core collapse.  However, the
specific processes involved in the production of black hole binaries \ilya{from isolated binaries}
could alter the spin \ilya{magnitude} distribution of those holes \ilya{relative to the X-ray binary systems}; for example, close
tidal interactions could spin up the core, or stripping of the
envelope could reduce the available angular momentum \citep{2016MNRAS.462..844K,Zaldarriaga:2017,HotokezakaPiran:2017}.

@@

\ilya{The spin directions in isolated binaries, whether evolving via the classical channel through a common-envelope phase \citep{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017} or through chemically homogeneous evolution \citep{MandeldeMink:2016,Marchant:2016} are generally expected to be preferentially aligned.  Despite  observed spin misalignments in both massive stellar binaries \citep{Albrecht:2009,2014ApJ...785...83A} and BH XRBs \citep{Orosz:2001,Martin:2008b,Martin:2008,MorningstarMiller:2014}, mass accretion and tidal interactions will tend to realign the binary.  On the other hand, a supernova natal kick (if any) can change the orbital plane and misalign the binary \citep{2000ApJ...541..319K,2013PhRvD..87j4028G}; the supernova can also tilt the spin angle, as in the double pulsar J0737-3039 \citep{2011ApJ...742...81F}; and a variety of uncertain processes, such as wind-driven mass loss and post-collapse fallback, can couple the spin magnitude and direction distributions, contrary to our simplified assumptions.}

\ilya{\citet{2017CQGra..34cLT01V} also studied the possibility of distinguishing aligned and isotropic angular distributions of \ac{BBH} spins, but concluded that several hundred sources would be required to adequately separate models which included both parallel and anti-parallel spins for 'aligned' population.  In contrast, and in agreement with the present study, \citet{Stevenson:2017spin} found that only $\sim 5$ observations of rapidly spinning black holes would be necessary to distinguish isotropic and aligned spin distributions, though tens of detections would be required for more nuanced admixture models.  Meanwhile, recent studies by \citet{Fishbach:2017,Gerosa:2017} focused on inference on spin magnitudes.}

\ilya{In summary}, the angular distribution of spins in \acp{BBH} formed through isolated binary evolution is uncertain; nevertheless, it is more probable that spins are \ilya{preferentially} aligned \sout{and non-zero} after evolution through this channel than not, while an isotropic distribution of \sout{non-zero} spins is the natural outcome of dynamical formation processes.  \ilya{Therefore, if the current observational trend for low $\chieff$ continues with future gravitational-wave observations, it will be possible to either confirm that \ac{BH} spins are sometimes large in magnitude and isotropic in direction, yielding a strong indication of a dynamical formation origin; or that \ac{BH} spins are overwhelmingly small in magnitude, yielding a notable contradiction with the claimed X-ray binary spin measurements, particularly those for high-mass X-ray binaries.}

@@

@@

If the true distribution is isotropic and follows this simple model,
and our measurement uncertainties on $\chieff$ are $\simeq 0.1$, then
each subsequent measurement contributes on average a factor of
$\sim 3$ to the overall odds.  After ten additional events, then, the
odds ratio becomes $5 \times 3^{10} \simeq 3 \times 10^{5}$, or
$4.6 \sigma$, consistent with the results of the more detailed
calculation described above.  If the true distribution of spins
becomes half as wide ($\chieff \in [-0.125, 0.125]$ for isotropic and
$\chieff \in [0, 0.25]$ for aligned spins), with the same
uncertainties, then the existing odds ratio becomes $1.08$, and each
subsequent event drawn from the isotropic distribution contributes on
average a factor of $1.6$.  In this case, after 10 additional events,
the odds ratio becomes $150$, or $2.7\sigma$.  With small spin
magnitudes, our angular resolving power vanishes.

>>>>

\begin{abstract}
  The first direct detections of \acp{GW} from merging \acp{BBH} open
  a unique window into the formation environment of massive stars and
  their compact remnants.  One promising signature of the formation
  environment is the angular distribution of the \ac{BH} spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin orientations whereas
  binaries formed from pairs of stars born together are more likely to
  have spins preferentially aligned with the binary orbit as a
  consequence of their joint evolution toward a \ac{BBH} system.  We
  consider existing \ac{GW} measurements of $\chieff$, the
  best-measured combination of spin parameters, in the three likely
  binary black hole detections GW150914, LVT151012, and GW151226.  If
  binary black hole spin magnitudes extend to high values, as is
  suggested by observations of black hole X-ray binaries, we show that
  the data already exhibit a $\OOneSigmaIsoAligned\sigma$
  ($\OOneOddsIsoAligned$ odds ratio) preference for an isotropic
  angular distribution.  By considering the effect of an additional 10
  detections drawn from the various models in the suite we show that
  if all observations come from a single population such an augmented
  data set would enable at least a $\OTwoSigmaIsoAlignedMin\sigma$
  ($\OTwoOddsIsoAlignedMin$ odds ratio) distinction between the
  isotropic and aligned models for the assumed spin magnitude
  distributions, and in most cases better than $5\sigma$
  ($2.9 \times 10^{-7}$ odds ratio).  The existing preference for
  either an isotropic spin distribution or low spin magntiudes for the
  observed systems will be confirmed (or overturned) confidently in
  the near future by subsequent observations.
\end{abstract}

@@

Small values of $\chieff$ as exhibited in these systems can result
from either intrinsically small spins or larger spins whose direction
is mis-aligned with the orbital plane of the binary (i.e.\ spin
vectors with small $z$-components).  Mis-alignment is capable of
producing \emph{negative} values of $\chieff$, however, while small,
aligned spins will always have $\chieff \geq 0$.  This difference
provides strong discriminating power between the two angular
distributions, even without good information about the magnitude
distribution; to the extent that data support negative $\chieff$ they
weigh heavily against aligned models.  To quantify the degree of
support for these two alternate explanations of small $\chieff$ values
in the merging \ac{BBH} population, we compared the Bayesian evidence
for various simple models of the spin population using the \ac{GW}
data set.

@@

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic, which explains why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``low'' magnitude distribution; the most favoured
model among those with an aligned distribution also has the ``low''
magnitude distribution.  The odds ratio between the aligned and
isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
\ac{BBH} systems having zero spin is ruled out by the GW151226
measurements, which bound $\chieff \gtrsim 0.2$ at 90\% credibility
\citep{2016PhRvL.116x1103A}.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the near future
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming an
additional 10 detections drawn from each of our six zero-parameter
models for the spin distribution, we find the odds ratios shown in
Figure \ref{fig:O2-predictions}.  We find that most scenarios with an
additional 10 detections allow the the correct angular distribution to
be distinguished with greater than $5\sigma$ ($2.9 \times 10^{-7}$
odds) credibility; and in the most pessisimistic case the distinction
is typically $\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$
odds ratio).  While such future detections should permit a confident
distinction between \emph{angular} distributions, we will remain very
uncertain about the \emph{magnitude} distribution until we have a much
larger number of observations.  In Figure \ref{fig:O2-predictions},
the odds ratio between different magnitude distributions with the same
angular distribution is much closer to one than the odds ratio between
angular distributions.

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.3$, then it is
no longer possible to resolve the negative spin with a small number of
observations.  As noted below, however, spins observed in X-ray
binaries are typically large.  Additionally, models which do not
permit \emph{some} spins with $\chieff \gtrsim 0.1$ are ruled out by
the GW151226 observations \citep{2016PhRvL.116x1103A}.

In order to perform our analysis we need to select models for the
distribution of the spin magnitudes of stellar-mass black holes.
Observational data for such model selection is sparse.
\citet{2015PhR...548....1M} give current estimates of the spin
parameters for stellar-mass black holes, obtained using disk
reflection and/or disk continuum methods.  Most of the systems studied
are low-mass X-ray binaries rather than the high-mass X-ray binaries
that are likely to be the progenitors of double black hole binaries.
In addition, there are substantial systematic errors that can
complicate either type of analysis (see \citealt{2015PhR...548....1M}
for a discussion), and selection effects could yield a biased
distribution. Nonetheless, if we take the reported spin magnitudes as
representative then we find that there is a preference for high spins;
for example, 14 of the 19 systems with reported spins have
dimensionless spin parameters in excess of 0.5.  It is usually argued
that the masses and spin parameters of stellar-mass black holes are
unlikely to be altered significantly by accretion (low-mass donors may
not have enough mass and high-mass donors have a very short phase in
which they transfer mass; a variant of this long-standing argument was
presented by \citet{1999MNRAS.305..654K}), but see
\citet{2003MNRAS.341..385P,2015ApJ...800...17F}.  Thus the current
spin parameters probably are close to their values upon core collapse.
However, the specific processes involved in the production of black
hole binaries from isolated binaries could alter the spin magnitude
distribution of those holes relative to the X-ray binary systems; for
example, close tidal interactions could spin up the core, or stripping
of the envelope could reduce the available angular momentum
\citep{2016MNRAS.462..844K,2017arXiv170200885Z,2017arXiv170203952H}.

@@

The spin directions in isolated binaries, whether evolving via the
classical channel through a common-envelope phase
\citep{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017}
or through chemically homogeneous evolution
\citep{MandeldeMink:2016,Marchant:2016} are generally expected to be
preferentially aligned.  Despite observed spin misalignments in both
massive stellar binaries \citep{Albrecht:2009,2014ApJ...785...83A} and
BH XRBs
\citep{Orosz:2001,Martin:2008b,Martin:2008,MorningstarMiller:2014},
mass accretion and tidal interactions will tend to realign the binary.
On the other hand, a supernova natal kick (if any) can change the
orbital plane and misalign the binary
\citep{2000ApJ...541..319K,2013PhRvD..87j4028G}; the supernova can
also tilt the spin angle, as in the double pulsar J0737-3039
\citep{2011ApJ...742...81F}; and a variety of uncertain processes,
such as wind-driven mass loss and post-collapse fallback, can couple
the spin magnitude and direction distributions, contrary to our
simplified assumptions.

\citet{2017CQGra..34cLT01V} also studied the possibility of
distinguishing aligned and isotropic angular distributions of \ac{BBH}
spins, but concluded that several hundred sources would be required to
adequately separate models which included both parallel and
anti-parallel spins for an ``aligned'' population.  In contrast, and
in agreement with the present study, \citet{Stevenson:2017spin} found
that only $\sim 5$ observations of rapidly spinning black holes would
be necessary to distinguish isotropic and aligned spin distributions,
though tens of detections would be required for more nuanced admixture
models.  Meanwhile, recent studies by
\citet{2017arXiv170306869F,2017arXiv170306223G} focused on inference
on spin magnitudes.

In summary, the angular distribution of spins in \acp{BBH} formed
through isolated binary evolution is uncertain; nevertheless, it is
more probable that spins are preferentially aligned after evolution
through this channel than not, while an isotropic distribution of
spins is the natural outcome of dynamical formation processes.
Therefore, if the current observational trend for low $\chieff$
continues with future gravitational-wave observations, it will be
possible to either confirm that \ac{BH} spins are sometimes large in
magnitude and isotropic in direction, yielding a strong indication of
a dynamical formation origin; or that \ac{BH} spins are overwhelmingly
small in magnitude, yielding a notable contradiction with the claimed
X-ray binary spin measurements, particularly those for high-mass X-ray
binaries.

@@

If the true distribution is isotropic and follows this simple model,
and our measurement uncertainties on $\chieff$ are $\simeq 0.1$, then
each subsequent measurement contributes on average a factor of
$\sim 3$ to the overall odds.  After ten additional events, then, the
odds ratio becomes $5 \times 3^{10} \simeq 3 \times 10^{5}$, or
$4.6 \sigma$, consistent with the results of the more detailed
calculation described above.  If the true distribution of spins
becomes half as wide ($\chieff \in [-0.125, 0.125]$ for isotropic and
$\chieff \in [0, 0.25]$ for aligned spins), with the same
uncertainties, then the existing odds ratio becomes $1.08$, and each
subsequent event drawn from the isotropic distribution contributes on
average a factor of $1.6$.  In this case, after 10 additional events,
the odds ratio becomes $150$, or $2.7\sigma$.  With small spin
magnitudes, our angular resolving power vanishes, as discussed in more
detail in Section \ref{sec:smallspins}.

@@
